---
title: "CSIT165 Initiative, Global Health Union (GHU) and Private Diagnostic Laboratories (PDL)."
author: "Haidee Hangis"
date: "2026-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r load-shape-merge, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(tidyr)
library(dplyr)

# Load data
confirmed  <- read.csv("time_series_covid19_confirmed_global.csv")
deaths     <- read.csv("time_series_covid19_deaths_global.csv")
recovered  <- read.csv("time_series_covid19_recovered_global.csv")

# Reshape
confirmed_long <- pivot_longer(confirmed, cols = starts_with("X"),
                               names_to = "date", values_to = "confirmed")

deaths_long <- pivot_longer(deaths, cols = starts_with("X"),
                            names_to = "date", values_to = "deaths")

recovered_long <- pivot_longer(recovered, cols = starts_with("X"),
                               names_to = "date", values_to = "recovered")

# Merge the three data sets into one
merged <- left_join(confirmed_long, deaths_long,
                    by = c("Province.State", "Country.Region", "Lat", "Long", "date"))

merged <- left_join(merged, recovered_long,
                    by = c("Province.State", "Country.Region", "Lat", "Long", "date"))

# Fix date
merged$date <- as.Date(gsub("X", "", merged$date), format = "%m.%d.%y")

```

## Objective 1. Origin of Covid-19 
>
Based on the maximum number of confirmed cases and deaths on the first recorded day in the dataset, China is the country of origin for Covid-19. China had 444 confirmed cases and 17 deaths registered in 2020-01-22. Being Hubei China a densely populated place is not surprise why the virus spread so rapidly in its early stages.
>

```{r}
first_day <- min(merged$date) #identifying the first recorded day in my merged dataset
first_day
first_day_data <- merged[merged$date == first_day, ] #filtering only the first day
max_confirm <- first_day_data[which.max(first_day_data$confirmed), ] # find the location with the most confirmed cases
max_confirm
max_deaths <- first_day_data[which.max(first_day_data$deaths), ] # find the location with the most confirmed deaths
max_deaths
#Comparing the country with the most confirmed cases and the country with the most deaths
if (max_confirm$Country.Region == max_deaths$Country.Region) {
  #origin <- max_confirm$Country.Region
  origin <- max_confirm
  print(paste("Predicted origin of COVID-19 is:", origin))
} else {
  origin <- max_confirm
  print("Confirmed and death origins do not match.")
}
origin
```

## Objetive 2. Most recent area with a first confirmed case
>
After putting aside all places with previous confirmed cases, the Province Pitcairn Islands in United Kingdom is the last region to report its first confirmed case in 2022-07-20.This happened two years and six months after the first case was confirmed in China.
>

```{r}

# Get all unique locations
locations <- unique(merged[, c("Country.Region", "Province.State")])

# Create empty results data frame
first_cases <- data.frame(
  Country = character(),
  Province = character(),
  Lat = numeric(),
  Long = numeric(),
  FirstCaseDate = as.Date(character())
)

# Loop through each location
for (i in 1:nrow(locations)) {
  
  country <- locations$Country.Region[i]
  province <- locations$Province.State[i]
  
  # Subset for this location
  subset_data <- merged[merged$Country.Region == country &
                        merged$Province.State == province, ]
  
  # Days with confirmed > 0
  positive_days <- subset_data[subset_data$confirmed > 0, ]
  
  # If this location ever had a case
  if (nrow(positive_days) > 0) {
    
    first_date <- min(positive_days$date)
    
    # Extract coordinates from the FIRST row of subset_data
    lat_val <- subset_data$Lat[1]
    long_val <- subset_data$Long[1]
    
    # Add row to results
    first_cases <- rbind(
      first_cases,
      data.frame(
        Country = country,
        Province = province,
        Lat = lat_val,
        Long = long_val,
        FirstCaseDate = first_date
      )
    )
  }
}

# Find the most recent first-case date
latest_date <- max(first_cases$FirstCaseDate)

# Extract the most recent location (one-row data frame)
most_recent_location <- first_cases[first_cases$FirstCaseDate == latest_date, ]

# Define 'recent' for Question 3
recent <- most_recent_location
recent

```

## Objective 3. How far away are the areas from objective 2 from where the first confirmed case(s) occurred
>
Pitcairn Islands in United Kingdom is 8746.97 miles away from Hubei, China where the first confirmed cases ocurred. The long distance between Hubei, China and Pitcairn Islands offers a logical explanation of why it took more than two years for the virus to spread onto this remote and more isolated place in the sense of its traveling connections are more limited.
>

```{r}
# Load geosphere
library(geosphere)

# Check that origin exists and is a data frame
if (!exists("origin") || !is.data.frame(origin) || nrow(origin) != 1) {
  stop("Error: 'origin' must be a one-row data frame created in Question 1.")
}

# Check that recent exists and is a data frame
if (!exists("recent") || !is.data.frame(recent) || nrow(recent) != 1) {
  stop("Error: 'recent' must be a one-row data frame created in Question 2.")
}

# Check that coordinates exist
if (is.null(origin$Lat) || is.null(origin$Long)) {
  stop("Error: origin does not contain Lat/Long columns.")
}

if (is.null(recent$Lat) || is.null(recent$Long)) {
  stop("Error: recent does not contain Lat/Long columns.")
}

# --- EXTRACT COORDINATES ---------------------------------------------------

origin_coords <- c(origin$Long, origin$Lat)
recent_coords <- c(recent$Long, recent$Lat)

# --- DISTANCE CALCULATION --------------------------------------------------

# Distance in meters
distance_meters <- distm(origin_coords, recent_coords)

# Convert meters to miles
distance_miles <- distance_meters * 0.000621371

# --- BUILD REGION NAMES ----------------------------------------------------

recent_region <- paste(recent$Province, recent$Country, sep = ", ")
origin_region <- paste(origin$Province.State, origin$Country.Region, sep = ", ")

# --- PRINT RESULT ----------------------------------------------------------

cat(recent_region, "is", round(distance_miles, 2),
    "miles away from", origin_region, "\n")

```
## Objective 4. Caracterizing diseases using risk scores.
>
The lowest‑risk area Jiansu, China has a very small ratio of deaths to confirmed cases, meaning outcomes there are relatively better.
The highest‑risk area North Korea has a much larger proportion of deaths relative to confirmed cases, indicating worse outcomes.
The global risk score 1.0073 provides a baseline for comparison.
>

```{r}
clean_data <- merged[!is.na(merged$Lat) & merged$Lat != 0 &
                     !is.na(merged$Long) & merged$Long != 0, ]
latest_day <- max(clean_data$date)
latest_data <- clean_data[clean_data$date == latest_day, ]
latest_data$risk_score <- ifelse(
  latest_data$confirmed > 0,
  100 * (latest_data$deaths / latest_data$confirmed),
  NA
)
min_risk <- min(latest_data$risk_score, na.rm = TRUE)

lowest_risk_data <- latest_data[latest_data$risk_score == min_risk, ]

# Break ties by choosing the one with most confirmations
lowest_risk <- lowest_risk_data[which.max(lowest_risk_data$confirmed), ]
max_risk <- max(latest_data$risk_score, na.rm = TRUE)

highest_risk_data <- latest_data[latest_data$risk_score == max_risk, ]

# Break ties by choosing the one with most confirmations
highest_risk <- highest_risk_data[which.max(highest_risk_data$confirmed), ]
global_confirmed <- sum(latest_data$confirmed, na.rm = TRUE)
global_deaths <- sum(latest_data$deaths, na.rm = TRUE)

global_risk <- 100 * (global_deaths / global_confirmed)
cat("Lowest-risk area:", lowest_risk$Province.State, ",", lowest_risk$Country.Region,
    "with risk score =", round(lowest_risk$risk_score, 4), "\n")

cat("Highest-risk area:", highest_risk$Province.State, ",", highest_risk$Country.Region,
    "with risk score =", round(highest_risk$risk_score, 4), "\n")

cat("Global risk score =", round(global_risk, 4), "\n")

```

## Objective 5. Two tables with the top 5 countries that have the most COVID-19 related confirmations and and deaths.
>
The tables below show that the top 5 countries with most covid-19 confirmed cases are:
>
1. USA
2. India
3. Brazil
4. France
5. Germany
>
And the top 5 countries with most covid-19 Deaths are:
>
1. USA
2. Brazil
3. India
4. Mexico
5. Russia
>
The first three countries in the list with most Covid-19 deaths have a correlation with the list of countries with most confirmed cases.
>

```{r}
countries <- unique(merged$Country.Region)
country_confirmations <- c()
country_deaths <- c()
for (c in countries) {
  
  # subset for this country
  subset_data <- merged[merged$Country.Region == c, ]
  
  # sum all confirmations and deaths
  total_confirmed <- sum(subset_data$confirmed, na.rm = TRUE)
  total_deaths <- sum(subset_data$deaths, na.rm = TRUE)
  
  # store results
  country_confirmations[c] <- total_confirmed
  country_deaths[c] <- total_deaths
}
confirm_df <- data.frame(
  Country = names(country_confirmations),
  Confirmations = as.numeric(country_confirmations)
)

deaths_df <- data.frame(
  Country = names(country_deaths),
  Deaths = as.numeric(country_deaths)
)
top5_confirm <- confirm_df[order(-confirm_df$Confirmations), ][1:5, ]
top5_deaths <- deaths_df[order(-deaths_df$Deaths), ][1:5, ]
library(knitr)
kable(top5_confirm, caption = "Top 5 Countries by Confirmations")
kable(top5_deaths, caption = "Top 5 Countries by Deaths")

```

